---
title: "Data Challenge for Natinal Data Managment Center"
author: "Haymanot B."
date: "2024-06-06"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1.  Preprocessing and EDA: Based on the given dataset(champs.csv) and the decoded variables in table 1, do the following preprocessing and Exploratory Data Analysis (EDA)

    **Ans:** First call or load necessary R packages from repository (CRAN)

```{r message=FALSE,warning=FALSE}
library(dplyr)
library(readr)
require(corrplot)
library(graphics)
library(grDevices)
library(caret)
library(kernlab)
library(randomForest)
library(gbm)
library(xgboost)
library(e1071)
library(pROC)
library(ggplot2)
library(pander)
library(pandoc)
library(tidyr)
```

A). read the dataset

**Ans:**

```{r message=FALSE, warning=FALSE}
library(readr)
df<- read_csv("CHAMPS.csv")
# View(CHAMPS)
```

B). How many rows and columns are they in the dataset?

**Ans:**

```{r message=FALSE, warning=FALSE}
dimensions <- dim(df)
num_rows <- dimensions[1]
num_cols <- dimensions[2]
dimensions
rows <- nrow(df)
rows
cols <- ncol(df)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
cat("The dataset has", rows, "rows and", cols, "columns.\n")
```

C). Enumerate the columns of the dataset

**Ans:**

```{r message=FALSE, warning=FALSE}
columns <- colnames(df)
#columns
```

D). Rename the columns. Example: rename column dp_013 to case_type

**Ans:** Lets select only few columns to show these renaming challenges

```{r message=FALSE, warning=FALSE}
df1 <- df %>% select(id_ver_nmb, champs_id, dp_013, dp_108, dp_118) %>% 
  rename(case_type = dp_013, underlying_cause = dp_108, maternal_condition = dp_118)
head(df1,4)
```

E). Rename values. Example: rename CH00716 to Stillbirth. Do the same for others too.

**Ans:** Lets check the levels of the variable exclusively

```{r message=FALSE, warning=FALSE}
table(df1$case_type)
 # Then using mutate we can recode as
df2 <- df1 %>%
   mutate(case_type = recode(case_type,
                             "CH00716" = "Stillbirth",
                             "CH01404" = "Death in the first 24 hours",
                             "CH01405" = "Early Neonate (1 to 6 days)",
                             "CH01406" = "Late Neonate (7 to 27 days)",
                             "CH00718" = "Infant (28 days to less than 12 months)",
                             "CH00719" = "Child (12 months to less than 60 months)"))
# df2
```

-   Rename error value Abruptio placenta under maternal_condition(dp_118) column

```{r message=FALSE, warning=FALSE}
  df2 <- df2 %>%
  mutate(maternal_condition = recode(maternal_condition,
                           "Abruptio placenta" = "Abruption placenta"))
```

F). Show the proportion of null values in each column.

**Ans**: Calculate the proportion of null values in each column and summarize the proportions with rounded values and display only the first 5 or 6 results

```{r message=FALSE, warning=FALSE}

pander(df2 %>%
  summarise(across(everything(), ~ sum(is.na(.)) / nrow(df2))) %>%
  gather(key = "variables", value = "Missing proportion") %>% head(10))
```

## 2. Descriptive Data analysis: looking back to the dataset above

A). What are the magnitude and proportion of each of the infant underlying cause for child death?

Lets first remove missing value and then we should sorted by count/magnitude of underlying_cause

```{r message=FALSE, warning=FALSE}
underlying_cause_counts <- df2 %>%
  filter(!is.na(underlying_cause)) %>%
  group_by(underlying_cause) %>%
  summarize(count = n()) %>%
  mutate(proportion = round((count / sum(count)) * 100, 3)) %>%
  arrange(desc(count)) 
pander(head(underlying_cause_counts, 4))
```

However the third top of proportion is the value of **undetermined** and is not underling cause while it is indicating missing values. So I strongly belive that I should remove it before going to any further analysis.

```{r message=FALSE, warning=FALSE}
# Remove the third row
underlying_cause_counts <- underlying_cause_counts[c(1, 2, 4:97), ]
pander(head(underlying_cause_counts, 4))
```

## B. What are the proportion and magnitude of the maternal factors contributing for child death?

-   To calculate proportion I first removed missing observation even not affected the result and show the first top count value by order counts in descending order.

```{r message=FALSE, warning=FALSE}
sort_maternal_condition_counts <- df2 %>%
  filter(!is.na(maternal_condition)) %>%
  group_by(maternal_condition) %>%
  summarize(count = n()) %>%
  mutate(proportion = count / sum(count)) %>%
  arrange(desc(count)) 
pander(head(sort_maternal_condition_counts, 4))
```

## C. What are the proportion of the child death by the case type

```{r message=FALSE, warning=FALSE}
chprop <- df2 %>%
  filter(!is.na(case_type)) %>%
  group_by(case_type) %>%
  summarize(count = n()) %>%
  mutate(proportion = round((count / sum(count)) * 100, 3)) %>%
  arrange(desc(count)) 
pander(head(chprop))
```

3.  Correlation analysis: Using correlation or Heat Maps, show how each of the infant under lying conditions and maternal factors are correlated to the top three causes of the child death identified above under 2(A)

to extract only top underlying cause and maternal factors

```{r message=FALSE, warning=FALSE}
top_3_causes <- underlying_cause_counts[1:3, ]
pander(top_3_causes)
top_factor<-sort_maternal_condition_counts[1:3,]
pander(top_factor)
top_causes_list <- top_3_causes$underlying_cause
pander(top_causes_list)
```

```{r message=FALSE, warning=FALSE}
df_top_causes <- df2 %>%
  filter(underlying_cause %in% top_causes_list)
df_top_causes_numeric <- df_top_causes %>%
  mutate(
    case_type = as.numeric(factor(case_type)),
    underlying_cause = as.numeric(factor(underlying_cause)),
    maternal_condition = as.numeric(factor(maternal_condition)))
```

## correlation analysis

```{r message=FALSE, warning=FALSE}
cor_matrix <- cor(df_top_causes_numeric[, c('case_type', 'underlying_cause', 'maternal_condition')], use = "complete.obs")

pander(print(cor_matrix))
```

```{r, message=FALSE, warning=FALSE, fig.width=12, fig.height=10}
library(gplots)
heatmap.2(cor_matrix,  col = cm.colors(256), margins = c(10, 10),   
          xlab = NULL,  ylab = NULL, main = "Heatmap Plot", trace = "none",        
          key = TRUE,   cexRow = 0.8,    cexCol = 0.8, key.title = "color key", labCol = F)       
```

## or

```{r message=FALSE, warning=FALSE}

corrplot(cor_matrix, method = "number", type = "lower", tl.cex = 0.7, tl.col = "blue",bg ="red" ,diag = TRUE,)
top_infant_causes<- head(underlying_cause_counts[order(-underlying_cause_counts$count),],6)
```



4.  Feature engineering:

<!-- -->

A.  Select the classification models *Logistic Regression, Support Vector Machine, AdaBoostClassifier, Random Forest Classifier , Gradient Boosting Classifier and XGBOOST* and train each on the dataset
B.  Import the appropriate package for each of the classification models above
C.  Rank the features based on their importance for each of the top underlying causes of child death identified above under 2(A), for each of the classification algorithms under (A )

**Ans:** Here what we do is select the best features (infant underlying causes and maternal factors) that contribute to the top three causes of child death that we identified in step 2A. Therefore let us select variables and remove missing values completely.


```{r message=FALSE, warning=FALSE}
# A. Import the appropriate package for each of the classification models
library(caret)
library(randomForest)
library(gbm)
library(xgboost)
library(e1071)
library(glmnet)
library(pROC)

# Prepare data
df_model <- df_top_causes_numeric %>%
  select(case_type, underlying_cause, maternal_condition) %>%
  na.omit()

df_model$case_type <- as.factor(df_model$case_type)
df_model$underlying_cause <- as.factor(df_model$underlying_cause)
df_model$maternal_condition <- as.factor(df_model$maternal_condition)

```

classification data as train data (70%) and test data (30%)
```{r message=FALSE, warning=FALSE}
set.seed(12345)
trainIndex <- createDataPartition(df_model$case_type, p = 0.7, list = FALSE, times = 1)
trainData <- df_model[trainIndex, ]
testData <- df_model[-trainIndex, ]
```



 Train classification models

```{r message=FALSE, warning=FALSE}
# Define control
control <- trainControl(method = "cv", number = 10)
models <- list()

# Logistic Regression
models$logistic_multinomial <- train(underlying_cause ~ .,  data = trainData,
  method = "multinom",  trControl = control)

# Support Vector Machine
models$svm <- train(underlying_cause ~ ., data = trainData, method = "svmRadial", trControl = control)

# Random Forest
models$random_forest <- train(underlying_cause ~ ., data = trainData, method = "rf", trControl = control)

# Gradient Boosting Classifier
models$gbm <- train(underlying_cause ~ ., data = trainData, method = "gbm", trControl = control, verbose = FALSE)

# XGBoost
models$xgboost <- train(underlying_cause ~ ., data = trainData, method = "xgbTree", trControl = control)
```


Rank features based on their importance

```{r, warning=FALSE, message=FALSE}
library(caret)
library(gbm)
library(xgboost)

# Function to get feature importance
get_feature_importance <- function(model) {
  if (inherits(model$finalModel, "randomForest")) {
    importance(model$finalModel)
  } else if (inherits(model$finalModel, "gbm")) {
    summary(model$finalModel, plotit = FALSE)$importance
  } else if (inherits(model$finalModel, "xgb.Booster")) {
    xgb.importance(model = model$finalModel)
  } else if (inherits(model$finalModel, "glmnet")) {
    coef(model$finalModel) # For glmnet (Lasso/Ridge) models
  } else {
    NULL # For models without feature importance
  }
}

# Get feature importance for each model
feature_importance <- lapply(models, get_feature_importance)

# Print feature importance
 pander(head(feature_importance))

```
    ------------------------------------------------------
           Feature           Gain      Cover    Frequency
    ---------------------- --------- --------- -----------
          case_type5        0.8226     0.659      0.58

          case_type3        0.1257    0.1224      0.14

     maternal_condition52   0.04242   0.1636      0.24

     maternal_condition32   0.00927   0.05498     0.04
    ------------------------------------------------------

5.  Model evaluation using the proper metrics

A. Import the appropriate evaluation metric packages

B. Using the appropriate n-fold cross validation and out of sample data, select the best preforming model from the candidate models under 4(A)

C. Ensemble the models and see the performance of the combination models on the data

D. Use Accuracy score metrics to evaluate the performance of the models above

E. Plot the AUC and ROC curve on the same graph to visualize and compare the performance of each of the models above

**Ans:**

```{r warning=FALSE,message=FALSE}
# Import evaluation metric packages
library(pROC)
library(caret)
# Predict and evaluate models
results <- lapply(models, predict, newdata = testData)
accuracy <- sapply(results, function(pred) {
  mean(pred == testData$underlying_cause)
})

# Print accuracy of individual models
pander(accuracy)

# Ensemble the models using majority voting
ensemble_pred <- apply(do.call(cbind, results), 1, function(row) {
  as.numeric(names(sort(table(row), decreasing = TRUE))[1])
})
ensemble_accuracy <- mean(ensemble_pred == testData$underlying_cause)
pander(ensemble_accuracy)
```

then we can ase Accuracy score metrics to evaluate the performance of the models above

```{r, warning=FALSE, message=FALSE}
accuracy_metrics <- sapply(results, function(pred) {
  confusionMatrix(pred, testData$underlying_cause)$overall['Accuracy']
})
panderOptions('table.split.table', 120)
pander(accuracy_metrics)
```


## Plot AUC and ROC curves

```{r message=FALSE, warning=FALSE}
roc_data <- lapply(results, function(pred) {
  roc(testData$underlying_cause, as.numeric(pred))
})
```

## Plot ROC curves

```{r message=FALSE, warning=FALSE, fig.height=10, fig.width=12}
par(mfrow = c(2, 3))
for (i in 1:length(roc_data)) {
  plot(roc_data[[i]], col = i + 1, main = paste("ROC Curve for", names(models)[i]))
}

```

6.  Result Visualization: Import the appropriate visualization package and:

A. Plot the feature importance in descending order for each of the models using horizontal bar chart

B. Plot the top five infant underlying causes of the child death

C. Plot the top five maternal factors contributing to the child death

D. Plot the child death based on the case types

```{r,message=FALSE, warning=FALSE, fig.height=12, fig.width=12}
# Plot feature importance for each model
par(mfrow = c(2, 3))
for (model_name in names(models)) {
  importance <- feature_importance[[model_name]]
  if (is.numeric(importance)) {
    barplot(sort(importance, decreasing = TRUE), horiz = TRUE,ylim = c(0, 15), main = paste("Feature Importance for", model_name))
  } else {
    print(paste())
  }
}
```

```{r message=FALSE, warning=FALSE, fig.height=10, fig.width=12}
varImpPlot(models$random_forest$finalModel, main = "Feature Importance for Random Forest")
```

## Plot top five infant underlying causes of child death

```{r message=FALSE, warning=FALSE}
top_infant_causes<-top_infant_causes[1:5,]
```

```{r message=FALSE, warning=FALSE, fig.width=10}
ggplot(top_infant_causes, aes(x = reorder(underlying_cause, -count) , y = count )) +
  geom_bar(stat = "identity", fill = 1:5 , colour = "red") +
  coord_flip() + 
  labs(title = "Top 5 Infant Underlying Causes of Child Death",
       x = "Underlying_Cause",
       y = "Count")
```

## Plot top five maternal factors contributing to child death

```{r message=FALSE, warning=FALSE}
top_maternal_factors <- head(sort_maternal_condition_counts[order(-sort_maternal_condition_counts$count), ], 6)
```

```{r message=FALSE, warning=FALSE}
top_maternal_factors<-top_maternal_factors[1:5,]
```

```{r message=FALSE, warning=FALSE}
top_maternal_factors<- top_maternal_factors %>%
  mutate(maternal_condition = recode(maternal_condition,
                                     "Fetus and newborn affected by other forms of placental separation and hemorrhage (Abruption placenta)" = 
                                     "(Abruption placenta)"))
```

```{r message=FALSE, warning=FALSE}
top_maternal_factors<- top_maternal_factors %>%
  mutate(maternal_condition = recode(maternal_condition, "Fetus and newborn affected by other forms of placental separation and hemorrhage" = "(F & N affected by PSH)"))
```

```{r message=FALSE, warning=FALSE, fig.dim=12}
library(ggplot2)
library(stringr)

ggplot(top_maternal_factors, aes(x = reorder(maternal_condition, -count), y = count)) +
  geom_bar(stat = "identity", fill = c("#00abcc","green", "blue", "red", "orange"), colour = "blue") +
  coord_flip() +
  labs(
    title = "Top 5 Maternal Factors Contributing to Child Death",
    x = "Maternal Condition",
    y = "Count"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10))

```

## Plot distribution of child deaths based on case types

```{r message=FALSE, warning=FALSE, fig.width=12}
ggplot(top_infant_causes, aes(x = reorder(underlying_cause, -count), y = count)) +
  geom_bar(stat = "identity" , fill = c("#00abcc","green", "blue", "red", "orange") , colour = "red") +
  coord_flip() +
  labs(title = "Distribution of Child Death by Case Type",
       x = "Case Type",
       y = "Count")
```
